Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-11-02 manuscript/manuscript.tex
Fixes in manuscript

Lithosphere's density variation with depth has been studied for almost a century. 
On the first years \citet{Athy1930} obtained a decreasing exponential relation between both quantities by studying shale samples, while in the following decades other density functions have been proposed for different rock types \citep[e.g.,][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, the density variation with depth has been taken into account in forward and inversion gravity models, mostly applied to grabens and basins \citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001, Welford2010}.

@@

The main issue that should be overcome is taking into account the curvature of the Earth, thus the forward model must be defined in spherical coordinates.
A common way to achieve this is to discretize the Earth in spherical prisms known as tesseroids, which are defined by pairs of latitude, longitude and radial boundaries (see Figure \ref{fig:tesseroid-uieda}).
Calculating the gravity fields generated by an arbitrary tesseroid on any external point involves the resolution of volume integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion \citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

In order to develop a forward model that computes the gravity field generated by any tesseroid with arbitrary variable density, the Taylor series expansion is not well suited, because it would need to derive the series expansion terms for each density function.
On the other hand, the GLQ consists in approximating the integral by a weighted sum of the effect of point masses located at scaled nodes of the Legendre polynomials, where the density information is gathered into these weights.
Thus, this kind of numerical approximation allows to calculate the gravity fields for any density function without any other information but the function itself.

@@

\citet{Ku1977} documented that the GLQ becomes less accurate when the computation point is closer to the tesseroid or for smaller GLQ order. 
Although the model developed by \citet{Uieda2016} uses a second order GLQ, it ensures the accuracy implementing a modified version of the adaptive discretization of \citet{Li2011}.
It consist in splitting the tesseroid into smaller ones when a certain distance-size ratio is lower than a predefined value ($D$) that controls the accuracy of the computation. 
\citet{Uieda2016} have also objectively obtained standard values of $D$ for gravity potential, gradients and tensor components comparing the fields generated by a spherical shell, which constitutes a special case of the volume integrals that has analytical solution \citep{LaFehr1991, Mikuska2006, Grombein2013}\todo{citar mas??}.

In addition, the varying density inside an arbitrary tesseroid introduces a new kind of error into the numerical approximation.
This encouraged us to develop a new density-based discretization algorithm that divides the tesseroid into smaller ones taking into account how the density function varies.
The number of subdivisions and thus the accuracy of the approximation will be controlled by a delta ratio $\delta$, a predefined value by the user.

We have enhanced the implementation of \citet{Uieda2016} with a new forward gravity model that allows us to compute the gravity fields generated by any tesseroid with an arbitrary density function on any external point.
It is based on the GLQ approximation, makes use of the modified distance-size discretization and the new density-based discretization algorithms that guarantee the accuracy even for high varying density functions.
It is written in Python and Cython languages, constituting an easy to use library that runs precompiled code for the more time consuming functions.
It also makes use of some previously existing classes and functions from Fatiando a Terra, what saved us time and allowed us to focus on the new enhancements.
Furthermore, we have the intention to include it into a future release of the Fatiando a Terra library in order to make its distribution and maintenance easier.

@@

In the same way as for an homogeneous density, we can observe from equation \ref{eq:glq-var-dens} that the GLQ approximates the gravity fields of a tesseroid as the ones generated by $N_\lambda N_\phi N_r$ point masses located at the nodes of the Legendre polynomials, where $N_i$ are the orders of the quadrature for each integration, with $i \in \{ \lambda, \phi, r \}$.
Nevertheless, in the variable density case, the information of the density function is summarised into the weights of the sum by taking into account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with well fitted discretization algorithms, approximates the gravity fields with good accuracy.\todo{no se si poner esta ultima oracion}

@@

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/spherical-shell.pdf}
\caption{
Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
The computation point $Q$ is located on the $z$ axe at a distance $r$ from the centre of the shell.
For practical purposes we will assume that $Q$ is outside the outer radii, i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity potential generated by a spherical shell with a variable density in depth on any outside point located on the $z$ axe.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the integral in equation \ref{eq:shell-pot} reproduces the gravity potential field on any outside point at distance $r$ from the its centre.

@@

Moreover, the new forward model is based on the preexisting code developed by \citet{Uieda2016} (and included in Fatiando a Terra v.0.5, \citet{Uieda2013}) that is able to compute the gravitational effects of tesseroids with homogeneous density.
On that implementation \citet{Uieda2016} have optimised the more time consuming functions through the Numba compiler.
In contrast, we have decided to develop our new library in Cython language due to difficulties when passing a Python function as argument to precompiled Numba methods, keeping the homogeneous tesseroid calculation and adding the new variable density computations.
The Cython language is a superset of Python language that allows to generate efficient C code through the Cython compiler, producing a library whose functions can be called from regular Python code.

It is also compatible with the preexisting open-source library Fatiando a Terra and its usage is very similar to the one for homogeneous tesseroids:

@@

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance between the geometric centre of the tesseroid and the computation point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for at least one spherical coordinate, then the tesseroid must be divided by half in that direction or directions.
A more complete description of the modified adaptive discretization algorithm can be found in \citet{Uieda2016}.
On the contrary, if the inequality is satisfied, the algorithm computes the desired gravity field using the GLQ approximation.

@@

On the other hand, the value assigned to the distance-size ratio $D$ cannot be easily related to the error of the approximation, thus the choice of the value of $D$ to get an acceptable accuracy must be objectively determined beforehand.
In order to overcome this, \citet{Uieda2016} compared the numerical computation with the analytical solution for a spherical shell with homogeneous density.

@@

Secondly, we must define a linear function $\rho_l(r')$ that assumes the same values of the normalised density $\rho_n(r')$ on the boundary radii:

@@

In order to determine the minimum value for the distance-size ratio $D$ such the numerical approximation achieves an acceptable accuracy we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, the second one represents a higher varying function, while the later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

@@

\noindent and the value of $b$ equal to the thickness of the spherical shell we will use to perform the test.
In this way, the density function assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum of the ones for a pure exponential function and for an homogeneous spherical shell.

We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, without applying the density-based discretization algorithm.
On both cases, the constant $b$ is equal to the thickness of the shell: 1km for the thin shell and 35km for the thick one.
The results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

@@

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/exponential-b-thin.pdf}
\caption{
    Results of the distance-size ratio determination for different exponential density functions in case of a spherical shell of 1km of thickness.
    (a) Density functions used in the tests, each one with a different value of $b$ and normalised in such way that they reaches a density of 2900kg/m$^{3}$ at the bottom of the shell.
    (b) Differences between the Marussi tensor component $g_{zz}$ generated by the numerical model and the analytical solution for each density function. Each computation was performed on the big grid described in Table \ref{tab:grids}, for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy for the corresponding value of $D$.
}
\label{fig:D-exp-power-thin}
\end{figure}

@@

On Figures \ref{fig:speed-comparison}a and \ref{fig:speed-comparison}b we can see the results of these comparison.
In the linear density case the density-based discretization algorithm is not applied because the inequation \ref{eq:delta-density} is never satisfied.
For this reason, the homogeneous density model is up to 5 times faster than the variable density one in this case.
On the other hand, the density-based discretization algorithm makes the computation slower: up to 20 times slower than the homogeneous density code.
Nevertheless, this increase in the computation time pays off with a guarantee on its accuracy.

@@

We have finally computed the gravity potential $V$, the gradient components $g_x$, $g_y$ and $g_z$, and the Marussi tensor components $g_{xy}$ and $g_{zz}$ on a computation grid of 159$\times$163 nodes with a 0.05$^\circ$ spacing on both longitude and latitude directions at a 50km height over the reference ellipsoid.
The resulting gravity fields can be seen in Figures \ref{fig:neuquen-basin}c-h.
We have excluded the other components of the Marussi tensor from the Figure, although they have also been computed and can be found in the repository.

@@

The distance-size ratio determinations have been made through the comparison of the numerical approximations with the analytical solutions for the a spherical shell with variable density, which had to be analytically obtained.
These tests have been made without applying the density-based discretization algorithm in order to isolate the former one.
From the wide range of possible density variations, we've chosen a linear, an exponential and a discontinuous density functions.
The comparisons over the first two functions have shown that values of $D$ equal to 1, 2 and 8 for the calculation of the gravity potential, its gradient and the Marussi tensor components, respectively, produce an accuracy of 0.1\% or better.
On the other hand, the algorithm on its own does not guarantee this amount of accuracy in case of highly varying exponential functions or discontinuous densities.

We have performed similar tests for the density-based discretization algorithm comparing the numerical model with the analytical solutions for a spherical shell with various exponential densities: from low varying to high varying functions.
Each comparison was made for different values of $\delta$, keeping the default values of $D$ previously obtained.
This tests have shown that a value of $\delta = 0.2$ guarantees an accuracy of 0.1\% or better on every case, while the same results can be obtained performing this tests with the discontinuous density function.

>>>>

Lithosphere's density variation with depth has been studied for almost a century. 
On the first years \citet{Athy1930} obtained a decreasing exponential relation between both quantities by studying shale samples, while in the following decades other density functions have been proposed for different rock types \citep[e.g.,][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, varying densities have been taken into account in forward and inversion gravity models, mostly applied to grabens and basins \citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001, Welford2010}.

@@

The main issue that should be overcome is to take into account the curvature of the Earth, thus the forward model must be defined in spherical coordinates.
A common way to achieve this is to discretize the Earth in spherical prisms known as tesseroids, which are defined by pairs of latitude, longitude and radial boundaries (see Figure \ref{fig:tesseroid-uieda}).
The calculation of the gravity fields generated by an arbitrary tesseroid on any external point involves the resolution of volume integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion \citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

In order to develop a forward model that computes the gravity field generated by any tesseroid with arbitrary variable density, the Taylor series expansion is not well suited, because it would need the series expansion terms for each density function.
On the other hand, the GLQ consists in approximating the integral by a weighted sum of the effect of point masses located at scaled nodes of the Legendre polynomials, where the density variation is summarised into the weights.
Thus, this kind of numerical approximation allows to calculate the gravity fields for any density function without any other information but the function itself.

@@

\citet{Ku1977} proved that the GLQ becomes less accurate when the computation point is closer to the tesseroid or for smaller GLQ order. 
Although the model developed by \citet{Uieda2016} uses a second order GLQ, it ensures the accuracy implementing a modified version of the adaptive discretization of \citet{Li2011}.
It consist in splitting the tesseroid into smaller ones when a certain distance-size ratio is lower than a predefined value ($D$) that controls the accuracy of the computation. 
\citet{Uieda2016} have also objectively obtained standard values of $D$ for gravity potential, gradients and tensor components comparing the numerical model with the fields generated by a spherical shell, which constitutes a special case of the volume integrals that has analytical solution \citep{LaFehr1991, Mikuska2006, Grombein2013}\todo{citar mas??}.

In addition, a varying density inside an arbitrary tesseroid introduces a new kind of error into the numerical approximation.
This encouraged us to develop a new density-based discretization algorithm that divides the tesseroid into smaller ones taking into account the variation of the density function.
The number of subdivisions and thus the accuracy of the approximation will be controlled by a delta ratio $\delta$, predefined by the user.

We have enhanced the implementation of \citet{Uieda2016} with a new forward gravity model that allows to compute the gravity fields generated by any tesseroid with an arbitrary density function on any external point.
It is based on the GLQ approximation, makes use of the modified distance-size discretization and the new density-based discretization algorithms that guarantee the accuracy even for high varying density functions.
It is written in Python and Cython languages, constituting an easy to use library that runs precompiled code for the more time consuming functions.
It also makes use of some previously existing classes and functions from Fatiando a Terra, what saved us time and allowed us to focus on the new enhancements.
Furthermore, we have the intention to include it into a future release of the Fatiando a Terra library in order to make its distribution and maintenance easier.

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid as the ones generated by $N_\lambda N_\phi N_r$ point masses located at the nodes of the Legendre polynomials, where $N_i$ are the orders of the quadrature for each integration, with $i \in \{ \lambda, \phi, r \}$, in the same way as the GLQ works in the homogeneous density case.
Nevertheless, in this new variable density scenario, the information of the density function is summarised into the weights by taking into account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with well suited discretization algorithms, approximates the gravity fields with good accuracy.\todo{no se si poner esta ultima oracion}

@@

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/spherical-shell.pdf}
\caption{
Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
The computation point $Q$ is located on the $z$ axe at a distance $r$ from the centre of the shell.
For our purposes we will assume that $Q$ is outside the outer radii, i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity potential generated by a spherical shell with a variable density in depth on any outside point located on the $z$ axe.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the expression in equation \ref{eq:shell-pot} reproduces the gravity potential field on any outside point at distance $r$ from the its centre.

@@

The new forward model is based on the preexisting code developed by \citet{Uieda2016} (and included in Fatiando a Terra v.0.5, \citet{Uieda2013}) that is able to compute the gravitational effects of tesseroids with homogeneous density.
On that implementation \citet{Uieda2016} have optimised the more time consuming functions through the Numba compiler.
In contrast, we have decided to develop our new library in Cython language due to difficulties when passing a Python function as argument to precompiled Numba methods, keeping the homogeneous tesseroid calculation and adding the new variable density computations.
The Cython language is a superset of Python language that allows to generate efficient C code through the Cython compiler, producing a library whose functions can be called from regular Python code.

Moreover, it is also compatible with the preexisting open-source library Fatiando a Terra and its usage is very similar to the one for homogeneous tesseroids:

@@

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance between the geometric centre of the tesseroid and the computation point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for at least one spherical coordinate, then the tesseroid must be divided by half in that direction.
A more complete description of the modified adaptive discretization algorithm can be found in \citet{Uieda2016}.
On the contrary, if the inequality is satisfied for every spherical coordinate, the algorithm computes the desired gravity field using the GLQ approximation.

@@

On the other hand, the value assigned to the distance-size ratio $D$ cannot be easily related to the error of the approximation, thus the choice of the value of $D$ to get an acceptable accuracy must be empirically determined beforehand.
In order to overcome this, \citet{Uieda2016} compared the numerical computation with the analytical solution for a spherical shell with homogeneous density.

@@

Secondly, we must define a linear function $\rho_l(r')$ that assumes the same values as the normalised density $\rho_n(r')$ on the boundary radii:

@@

In order to determine the minimum value for the distance-size ratio $D$ such the numerical approximation achieves an acceptable accuracy, we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, the second one represents a higher varying function, while the later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

@@

\noindent and the value of $b$ equal to the thickness of the spherical shell we will use to perform the test.
In this way, the density function assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum of the ones for a pure exponential function (like the one in equation \ref{eq:density-exp}) and for an homogeneous spherical shell.

We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, without applying the density-based discretization algorithm.
On both cases, the constant $b$ is equal to the thickness of the shell: 1km for the thin shell and 35km for the thick one.
The results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, we exclude the ones corresponding to $g_{xx}$ and $g_{yy}$ due to their similarity with the ones obtained for $g_{zz}$, although they can be found in the repository.

@@

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/exponential-b-thin.pdf}
\caption{
    Results of the distance-size ratio determination for different exponential density functions in case of a spherical shell of 1km of thickness.
    (a) Density functions used in the tests, each one with a different value of $b$.
    (b) Differences between the Marussi tensor component $g_{zz}$ generated by the numerical model and the analytical solution for each density function. Each computation was performed on the big grid described in Table \ref{tab:grids}, for different values of the distance-size ratio $D$. If the difference is less than 0.1\%, we consider that the model has achieved an acceptable accuracy.
}
\label{fig:D-exp-power-thin}
\end{figure}

@@

On Figures \ref{fig:speed-comparison}a and \ref{fig:speed-comparison}b we can see the results of these comparison.
In the linear density case the density-based discretization algorithm is not applied because the inequation \ref{eq:delta-density} is never satisfied.
For this reason, the homogeneous density model is up to 5 times faster than the variable density one in this case.
On the other hand, the density-based discretization algorithm makes the computation slower: up to 20 times slower than the homogeneous density code.
Nevertheless, this increase in the computation time pays off with a guarantee\todo{guarantee is a noun?} on its accuracy.

@@

We have finally computed the gravity potential $V$, the gradient components $g_x$, $g_y$ and $g_z$, and the Marussi tensor components $g_{xy}$ and $g_{zz}$ on a computation grid of 159$\times$163 nodes with a 0.05$^\circ$ spacing on both longitude and latitude directions at a 50km height over the reference ellipsoid.
The resulting gravity fields can be seen in Figures \ref{fig:neuquen-basin}c-h.
We have excluded the other components of the Marussi tensor from the Figure, although they can be found in the repository.

@@

The distance-size ratio determinations have been made through the comparison of the numerical approximations with the analytical solutions for a spherical shell with variable density, which had to be analytically obtained.
These tests have been made without applying the density-based discretization algorithm in order to isolate the former one.
From the wide range of possible density variations, we've chosen a linear, an exponential and a discontinuous density functions.
The comparisons over the first two functions have shown that values of $D$ equal to 1, 2 and 8 for the calculation of the gravity potential, its gradient and the Marussi tensor components, respectively, produce an accuracy of 0.1\% or better.
On the other hand, the algorithm on its own does not guarantee this amount of accuracy in case of highly varying exponential functions or discontinuous densities.

We have performed similar tests for the density-based discretization algorithm comparing the numerical model with the analytical solutions for a spherical shell with various exponential densities: from low to high varying functions.
Each comparison was made for different values of $\delta$, keeping the default values of $D$ previously obtained.
This tests have shown that a value of $\delta = 0.2$ guarantees an accuracy of 0.1\% or better on every case, while the same results can be obtained performing this tests with the discontinuous density function.
